Characterizing the Splice Map of Turkey Hemorrhagic Enteritis Virus

### 2024-05-09 manuscript_thev_transcriptome.tex
second round of edits after jvi review

  pdftitle={Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus in a Turkey B-cell Line},

@@

\title{Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis
Virus in a Turkey B-cell Line}
\author{}
\date{\vspace{-2.5em}}

@@

\textbf{Running Title:} Novel Insights into Turkey Hemorrhagic Enteritis
Virus Transcriptome

\vspace{10mm}

Abraham Quaye\textsuperscript{1*}, Brett E. Pickett\textsuperscript{*},
Joel S. Griffitts\textsuperscript{*}, Bradford K.
Berges\textsuperscript{*}, Brian D.
Poole\({^\dagger}\)\textsuperscript{*}

@@

\textsuperscript{*}Department of Microbiology and Molecular Biology,
Brigham Young University\\
\textsuperscript{1}First-author\\
\({^\dagger}\) Corresponding Author

@@

A myriad of studies have elucidated the AdV transcriptome in fine detail
(17, 18). However, a large preponderance of studies focus on MAdVs --
specifically human AdVs. Thus, most of the current AdV gene expression
and replication knowledge is based on MAdV studies, which is generalized
for all other AdVs (10, 19). MAdV transcription is temporally regulated;
their genes are categorized into five early transcription units (E1A,
E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and
one major late transcription unit (MLTU or major late promoter {[}MLP{]}
region), which generates five families of late mRNAs (L1-L5) based on
the polyadenylation site. An additional gene (UXP or U exon) is located
on the reverse strand. The early genes encode non-structural proteins
such as enzymes or host-cell modulating proteins, primarily involved in
DNA replication, or providing the necessary intracellular niche for
optimal replication while late genes encode structural proteins that act
as capsid proteins, promote virion assembly, or direct genome packaging.
The immediate early genes E1A are expressed first, followed by the
delayed early genes, E1B, E2, E3 and E4. Then the intermediate early
genes, IVa2 and pIX are expressed followed by the late genes (10, 17,
18). It is noteworthy that the MLP shows basal transcriptional activity
during early infection (before DNA replication), with a comparable
efficiency to other early viral promoters, but it reaches its maximal
activity during late infection (after DNA replication). However, during
early infection only a subset of the MLP-derived transcripts are
expressed (10). MAdV makes an extensive use of alternative RNA splicing
to produce a very complex array of mRNAs. All but the pIX mRNA undergo
at least one splicing event. For instance, the MLTU produces over 20
distinct splice variants all containing three non-coding exons at the
5'-end (collectively known as the tripartite leader; TPL) (17, 18).
There is also an alternate three-exon 5' non-coding leader sequence
present in varying amounts on a subset of MLTU mRNAs (known as the x-,
y-, and z-leaders). Lastly, there is the i-leader exon, which is
infrequently included between the second and third TPL exons, and codes
for the i-leader protein (20). Thus, the MLTU produces a complex
repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced
onto different 3' coding exons which are grouped into five different
3'-end classes (L1-L5) based on polyadenylation site. Each transcription
unit (TU) contains its own promoter driving the expression of the array
of mRNA transcripts produced via alternative splicing in the unit (10,
17, 18). The promoters are activated at different phases of the
infection by proteins from previously activated TUs. Paradoxically, the
early-to-late phase transition during infection requires the L4 gene
products, 22K and 33K, which should only be available after the
transition. However, a promoter in the L4 region (L4P) that directs the
expression of these two proteins independent of the MLP was found,
resolving the paradox (10, 17, 21). During translation of AdV mRNA,
recent studies using long-read direct RNA sequencing strongly suggest
the potential usage of secondary start codons; adding to what was
already a highly complex system for gene expression (17, 22).

@@

\textbf{Early Region 2 (E2) and Intermediate Region (IM) transcripts}\\
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and
E2B and encodes three classical AdV proteins -- pTP and Ad-pol (E2B
proteins), and DBP (E2A protein) -- essential for genome replication
(17, 18). Unlike MAdV where two promoters are known (17), we discovered
only a single TSS (E2 TSS; 18,751 bp) for both E2A and E2B transcripts
in THEV. However, similar to MAdVs, E2A and E2B transcripts have
distinct TTSs, with E2B transcripts sharing the TTS of the IVa2
transcript of the IM region (17, 18) (\textbf{Figure 7}).

@@

The E2B region transcripts also start with the E2-5'UTR but extend
downstream to reach the TTS at 2334 bp in the IM region, which is in the
immediate context of a canonical PAS (position 2333 bp) where
polyadenylation likely occurs. The E2B transcripts, TRXPT\_6 and
TRXPT\_7, are almost identical except for an extra splice junction at
the 3'-end of TRXPT\_6 (\textbf{Figure 7}). TRXPT\_7 has the CP for both
classical proteins (pTP and Ad-pol) encoded in this region, with the pTP
ORF predicted to be spliced from two exons. The predicted splice
junction of pTP is corroborated by our data but the full transcript is
markedly longer than the predicted ORF, although, the encoded product
(pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of
pTP with secondary SC (secSC) usage. The CP of TRXPT\_6 slightly differs
from TRXPT\_7 because a new STC resulting from the extra splice site
forms a minimal truncation of the Ad-pol encoded from its secSC.

@@

\textbf{Early Region 3 (E3) transcripts}\\
The E3 region, nested within the MLTU, encodes proteins that modulate
and evade host immune defenses. In MAdVs, this region contains seven
ORFs expressed from multiple transcripts sharing the same TSS (from the
E3 promoter) but having different TTSs (10, 17, 18). However, some E3
transcripts use the TSS of the MLP. Due to sharing the same TSS, in
MAdVs, secSC usage is heavily relied on for gene expression in this
region as utilizing only the first SC cannot produce the downstream
transcripts in this TU. The 12.5K ORF and transcripts using the MLP TSS
are exceptions (17).

@@

As the splice junctions of TRXPT\_22, TRXPT\_23, TRXPT\_24, and
TRXPT\_29 share the same genomic space, their validation was done with a
single primer pair, and they were differentiated from each other by
cloning the cDNA and Sanger sequencing (\textbf{Supplementary PCR
methods}). In addition to corroborating the splice junctions for the
aforementioned transcripts, the Sanger sequencing results also showed a
distinct splice variant undetected in our RNA-seq transcriptome. This
was a three-exon transcript (TRXPT\_30) with identical first and last
exons as TRXPT\_23, but also contained the second exon of TRXPT\_24
(\textbf{Figure 8}). TRXPT\_30 encodes a novel 140 residue, 15.7kDa
protein (e22K), spanning all three exons. Interestingly, the last 81
C-terminus residues of e22K are identical to 22K (89 residues), a
single-exon ORF predicted to use the same SC as 33K. Just as seen for
33K, the first 20 bp of 22K is intronic, excluding the first 7 residues
of 22K from e22K. We consider e22K as a long variant of the predicted
22K ORF. Assuming TRXPT\_30 shares the same TSS and TTS as TRXPT\_23, it
would also have the downstream CP of TRXPT\_23.

@@

\textbf{Early Region 4 (E4) transcripts} This TU is found at the 3'-end
of the genome and expressed on the anti-sense strand. Based on
nucleotide position, ORF7 and ORF8 were predicted in this region (1);
however, as ORF7 is neither on the anti-sense strand nor transcribed
from a promoter in the E4 region; hence, we only classify ORF8 in this
TU. This is corroborated by our RNA-seq data, showing only one
transcript in this region on the anti-sense strand (\textbf{Figure 9}).
The transcript (TRXPT\_28) spans 25192-26247 bp and is spliced at
25701-26055 bp, forming a two-exon transcript. The second exon fully
matches the predicted ORF8 with 12 extra base pairs at the 3'-end.
However, we identified a SC 192 bp upstream of the predicted SC in the
first exon from which an in-frame protein is encoded. We consider this
longer isoform (eORF8 -- 26.4 kDa, 229 residues) as the genuinely
expressed ORF. We also identified a canonical PAS 11 bp upstream of TTS
(\textbf{Supplementary Table S2}).

@@

We identified five TTSs (10,549, 12,709, 16,870, 22,116, 25,168 bp) in
this TU, which we consider as corresponding to the five late mRNA
classes (L1-L5), respectively. L1 mRNAs include TRXPT\_8, encoding the
52K ORF as predicted. L2 mRNAs include TRXPT\_16, TRXPT\_17, and
TRXPT\_18, all containing the sTPL with their respective coding exons.
They encode pIIIa, III (penton), and pVII, respectively. The L3 mRNAs,
TRXPT\_14 and TRXPT\_20, both encode the hexon (II) ORF but hexon is the
only ORF encoded on TRXPT\_14, whereas TRXPT\_20 encodes pX (pre-Mu),
pVI, and hexon in that order. L4 mRNAs, TRXPT\_9, TRXPT\_10, TRXPT\_11,
and TRXPT\_13 are the largest transcripts in the transcriptome and
encode several similar late proteins. TRXPT\_9 and TRXPT\_10 are very
similar but not identical. The last exon of TRXPT\_9 seems to be
truncated and likely shares the same TTS as TRXPT\_10. They both encode
pVII as the 5'-most ORF and also have the CP for pX, pVI, hexon, a
longer variant of protease (eProt) from an upstream in-frame SC, and
ORF12 (a novel 120 residue protein). Additionally they have the CP for
pVIII and eE3. TRXPT\_11 encodes hexon as its 5'-most ORF and also has
the CP for eProt, ORF12, e33K, pVIII and eE3. Typically, MLTU
transcripts splice the TPL onto a splice site just upstream of the ORF
to be expressed (17). While this holds true for most MLTU ORFs, several
late ORFs (pVI, protease, and ORF7) do not have such close proximity
splicing but are contained in larger transcripts such as these L4 mRNAs,
strongly suggesting the use of non-standard ribosomal initiation
mechanisms such as secSC usage or ribosome shunting described in other
AdVs for their translation (17, 29). TRXPT\_13, an E3 ORF utilizing the
MLP TSS, encodes the classical L4P genes, pVIII and eE3. Lastly, the L5
class transcript, TRXPT\_12, encodes Fiber as its 5'-most ORF but also
has the CP for ORF7. Interestingly, the CP of TRXPT\_12 and TRXPT\_27 of
the E3 TU are identical but are initiated from different TSSs. \newpage

\subsection{DISCUSSION/CONCLUSIONS}\label{discussionconclusions}

@@

AdV alternative splicing undergoes a regulated temporal shift in splice
site usage, previously thought to be limited to certain TUs. However,
recent studies suggest that AdVs routinely produce different
combinations of splice acceptor--donor pairs across all TUs (10, 17, 22,
30). The details of this phenomenon have been best studied for the E1A
and L1 units. AdVs modulate the activities of the splicing factor U2AF
and the cellular SR family of splicing factors (reviewed here (30)), and
encode several proteins that influence the RNA splice site used. This
phenomenon appears to occur in the THEV transcriptome, as the stringency
of splice acceptor-donor pairs selected decreases from the onset of the
late phase (\textbf{Figure 5}). Recent studies show that a virtually
unlimited number of combinatorial alternative splicing events occur in
an AdV lytic infection, resulting in a variety of novel transcripts (17,
22). It is unlikely that the entire repertoire of mRNA produced via this
mechanism will actually be translated. However, it has been speculated
that the plasticity in alternative RNA splicing enables AdVs to
fine-tune protein synthesis by providing different alternatively spliced
variants encoding the same protein under changing conditions, confering
an evolutionary advantage (17, 22).

@@

\subsection{MATERIALS AND METHODS}\label{materials-and-methods}

>>>>

  pdftitle={Characterizing the Splice Map of Turkey Hemorrhagic Enteritis Virus},

@@

\title{Characterizing the Splice Map of Turkey Hemorrhagic Enteritis
Virus}
\author{}
\date{\vspace{-2.5em}}

@@

Abraham Quaye\({^\dagger}\)\textsuperscript{,a}, Brett E.
Pickett\textsuperscript{a}, Joel S. Griffitts\textsuperscript{a},
Bradford K. Berges\textsuperscript{a}, Brian D.
Poole\textsuperscript{a,*}

@@

\textsuperscript{a}Department of Microbiology and Molecular Biology,
Brigham Young University\\
\({^\dagger}\)First-author\\
\textsuperscript{*}Corresponding Author

@@

A myriad of studies have elucidated the AdV transcriptome in fine detail
(17, 18). However, a large preponderance of studies focus on MAdVs --
specifically human AdVs. Thus, most of the current AdV gene expression
and replication knowledge is based on MAdV studies, which is generalized
for all other AdVs (10, 19). MAdV transcription is temporally regulated;
their genes are categorized into five early transcription units (E1A,
E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and
one major late transcription unit (MLTU or major late promoter {[}MLP{]}
region), which generates five families of late mRNAs (L1-L5) based on
the polyadenylation site. An additional gene (UXP or U exon) is located
on the reverse strand. The early genes encode non-structural proteins
such as enzymes or host-cell modulating proteins, primarily involved in
DNA replication, or providing the necessary intracellular niche for
optimal replication while late genes encode structural proteins that act
as capsid proteins, promote virion assembly, or direct genome packaging.
The immediate early genes E1A are expressed first, followed by the
delayed early genes, E1B, E2, E3 and E4. Then the intermediate early
genes, IVa2 and pIX are expressed followed by the late genes (10, 17,
18). It is noteworthy that the MLP shows basal transcriptional activity
during early infection (before DNA replication), with a comparable
efficiency to other early viral promoters, but it reaches its maximal
activity during late infection (after DNA replication). However, during
early infection only a subset of the MLP-derived transcripts are
expressed (10). MAdV makes an extensive use of alternative RNA splicing
to produce a very complex array of mRNAs. All but the pIX mRNA undergo
at least one splicing event. For instance, the MLTU produces over 20
distinct splice variants all containing three non-coding exons at the
5'-end (collectively known as the tripartite leader; TPL) (17, 18).
There is also an alternate three-exon 5' non-coding leader sequence
present in varying amounts on a subset of MLTU mRNAs (known as the x-,
y-, and z-leaders). Lastly, there is the i-leader exon, which is
infrequently included between the second and third TPL exons, and codes
for the i-leader protein (20). Thus, the MLTU produces a complex
repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced
onto different 3' coding exons which are grouped into five different
3'-end classes (L1-L5) based on polyadenylation site. Each transcription
unit (TU) contains its own promoter driving the expression of the array
of mRNA transcripts produced via alternative splicing in the unit (10,
17, 18). The promoters are activated at different phases of infection by
proteins from previously activated TUs. Paradoxically, the early-to-late
phase transition during infection requires the L4 gene products, 22K and
33K, which should only be available after the transition. However, a
promoter in the L4 region (L4P) that directs the expression of these two
proteins independent of the MLP was found, resolving the paradox (10,
17, 21). During translation of AdV mRNA, recent studies using long-read
direct RNA sequencing strongly suggest the potential usage of secondary
start codons; adding to what was already a highly complex system for
gene expression (17, 22).

@@

\textbf{Early Region 2 (E2) and Intermediate Region (IM) transcripts}\\
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and
E2B and encodes three classical AdV proteins -- pTP and Ad-pol (E2B
proteins), and DBP (E2A protein) -- essential for genome replication
(17, 18). Unlike MAdV where two promoters are known (17), we discovered
only a single TSS (E2 TSS; 18,751 bp) for both E2A and E2B transcripts
in THEV. However, E2A and E2B transcripts have distinct TTSs, with E2B
transcripts sharing the TTS of the IVa2 transcript of the IM region
similar to MAdVs (17, 18) (\textbf{Figure 7}).

@@

The E2B region transcripts also start with the E2-5'UTR but extend
downstream to reach the TTS at 2334 bp in the IM region, which is in the
immediate context of a canonical PAS (position 2333 bp) where
polyadenylation likely occurs. The E2B transcripts, TRXPT\_6 and
TRXPT\_7, are almost identical except for an extra splice junction at
the 3'-end of TRXPT\_6 (\textbf{Figure 7}). TRXPT\_7 has the CP for both
classical proteins (pTP and Ad-pol) encoded in this region, with the pTP
ORF predicted to be spliced from two exons. The predicted splice
junction of pTP is corroborated by our data but the full transcript is
markedly longer than the predicted ORF, although the encoded product
(pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of
pTP with secondary start codon (secSC) usage. The CP of TRXPT\_6
slightly differs from TRXPT\_7 because a new STC resulting from the
extra splice site forms a minimal truncation of the Ad-pol encoded from
its secSC.

@@

\textbf{Early Region 3 (E3) transcripts}\\
The E3 region, nested within the MLTU, encodes proteins that modulate
and evade host immune defenses. In MAdVs, this region contains seven
ORFs expressed from multiple transcripts sharing the same TSS (from the
E3 promoter) but having different TTSs (10, 17, 18). However, some E3
transcripts use the TSS of the MLP. Due to sharing the same TSS, in
MAdVs, secSC usage is heavily relied on for gene expression in this
region as utilizing only the first SC cannot produce the downstream
proteins in this TU. The 12.5K ORF and transcripts using the MLP TSS are
exceptions (17).

@@

As the splice junctions of TRXPT\_22, TRXPT\_23, TRXPT\_24, and
TRXPT\_29 share the same genomic space, their validation was done with a
single primer pair, and they were differentiated from each other by
cloning the cDNA and Sanger sequencing (\textbf{Supplementary PCR
methods}). In addition to corroborating the splice junctions for the
aforementioned transcripts, the Sanger sequencing results also showed a
distinct splice variant undetected in our RNA-seq transcriptome. This
was a three-exon transcript (TRXPT\_30) with identical first and last
exons as TRXPT\_23, which also contained the second exon of TRXPT\_24
(\textbf{Figure 8}). TRXPT\_30 encodes a novel 140 residue, 15.7kDa
protein (e22K), spanning all three exons. Interestingly, the last 81
C-terminus residues of e22K are identical to 22K (89 residues), a
single-exon ORF predicted to use the same SC as 33K. Just as seen for
33K, the first 20 bp of 22K is intronic, excluding the first 7 residues
of 22K from e22K. We consider e22K as a long variant of the predicted
22K ORF. Assuming TRXPT\_30 shares the same TSS and TTS as TRXPT\_23, it
would also have the downstream CP of TRXPT\_23.

@@

\textbf{Early Region 4 (E4) transcripts} This TU is found at the 3'-end
of the genome and expressed on the anti-sense strand. Based on
nucleotide position, ORF7 and ORF8 were predicted in this region (1);
however, as ORF7 is neither on the anti-sense strand nor transcribed
from a promoter in the E4 region, we only classify ORF8 in this TU. This
is corroborated by our RNA-seq data, showing only one transcript in this
region on the anti-sense strand (\textbf{Figure 9}). The transcript
(TRXPT\_28) spans 25192-26247 bp and is spliced at 25701-26055 bp,
forming a two-exon transcript. The second exon fully matches the
predicted ORF8 with 12 extra base pairs at the 3'-end. However, we
identified a SC 192 bp upstream of the predicted SC in the first exon
from which an in-frame protein is encoded. We consider this longer
isoform (eORF8 -- 26.4 kDa, 229 residues) as the genuinely expressed
ORF. We also identified a canonical PAS 11 bp upstream of TTS
(\textbf{Supplementary Table S2}).

@@

We identified five TTSs (10,549, 12,709, 16,870, 22,116, 25,168 bp) in
this TU, which we consider as corresponding to the five late mRNA
classes (L1-L5), respectively. L1 mRNAs include TRXPT\_8, encoding the
52K ORF as predicted. L2 mRNAs include TRXPT\_16, TRXPT\_17, and
TRXPT\_18, all containing the sTPL with their respective coding exons.
They encode pIIIa, III (penton), and pVII, respectively. The L3 mRNAs,
TRXPT\_14 and TRXPT\_20, both encode the hexon (II) ORF but hexon is the
only ORF encoded on TRXPT\_14, whereas TRXPT\_20 encodes pX (pre-Mu),
pVI, and hexon in that order. L4 mRNAs, TRXPT\_9, TRXPT\_10, TRXPT\_11,
and TRXPT\_13 are the largest transcripts in the transcriptome and
encode several similar late proteins. TRXPT\_9 and TRXPT\_10 are very
similar but not identical. The last exon of TRXPT\_9 seems to be
truncated and likely shares the same TTS as TRXPT\_10. They both encode
pVII as the 5'-most ORF and also have the CP for pX, pVI, hexon, a
longer variant of protease (eProt) from an upstream in-frame SC, and
ORF12 (a novel 120 residue protein). Additionally, they have the CP for
pVIII and eE3. TRXPT\_11 encodes hexon as its 5'-most ORF and also has
the CP for eProt, ORF12, e33K, pVIII and eE3. Typically, MLTU
transcripts splice the TPL onto a splice site just upstream of the ORF
to be expressed (17). While this holds true for most MLTU ORFs, several
late ORFs (pVI, protease, and ORF7) do not have such close proximity
splicing but are contained in larger transcripts such as these L4 mRNAs,
strongly suggesting the use of non-standard ribosomal initiation
mechanisms such as secSC usage or ribosome shunting described in other
AdVs for their translation (17, 29). TRXPT\_13, an E3 ORF utilizing the
MLP TSS, encodes the classical L4P genes, pVIII and eE3. Lastly, the L5
class transcript, TRXPT\_12, encodes Fiber as its 5'-most ORF but also
has the CP for ORF7. Interestingly, the CP of TRXPT\_12 and TRXPT\_27 of
the E3 TU are identical but are initiated from different TSSs. \newpage

\subsection{DISCUSSION}\label{discussion}

@@

AdV alternative splicing undergoes a regulated temporal shift in splice
site usage, previously thought to be limited to certain TUs. However,
recent studies suggest that AdVs routinely produce different
combinations of splice acceptor--donor pairs across all TUs (10, 17, 22,
30). The details of this phenomenon have been best studied for the E1A
and L1 units. AdVs modulate the activities of the splicing factor U2AF
and the cellular SR family of splicing factors (reviewed here (30)), and
encode several proteins that influence the RNA splice site used. This
phenomenon appears to occur in the THEV transcriptome, as the stringency
of splice acceptor-donor pairs selected decreases from the onset of the
late phase (\textbf{Figure 5}). Recent studies show that a virtually
unlimited number of combinatorial alternative splicing events occur in
an AdV lytic infection, resulting in a variety of novel transcripts (17,
22). It is unlikely that the entire repertoire of mRNA produced via this
mechanism will actually be translated. However, it has been speculated
that the plasticity in alternative RNA splicing enables AdVs to
fine-tune protein synthesis by providing different alternatively spliced
variants encoding the same protein under changing conditions, conferring
an evolutionary advantage (17, 22).

@@

\subsection{METHODS}\label{methods}
